Combining living and fossil taxa into phylogenies: the missing data issue

### 2014-12-01 Manuscript/TEM_draft.tex
Updated discussion plan + finished results

-Overall missing data is doesn't affect topological recovery in Bayesian (around minimum 70\% of topological topology in the worth scenario).

@@

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014}. Missing molecular data has been seen by some authors as an issue because it can, in some part of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing to infer topology and branch length)% NC: define phylogenetic signal here. 
    , especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by:
(i) including a "modest" number of highly-covered genes (i.e. approximatively of the genes that are coded for most of the species; \citealt{rouresite-specific2011}) %NC: what do you mean by highly-covered?
(ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citeauthor{sansomfossilization2013} \citeyear{sansomfossilization2013} but see \citeauthor{pattinsonphylogeny2014} \citeyear{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{wiensmissing2003}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold were it starts fucking up the phylogeny) but more the fact that Bayesian fails.

@@

We inferred the morphological matrix using the R package ape v3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (See \hyperref[supplementaries]{supplementaries}). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to an other \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters (i.e. $\alpha$) to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014}.

@@

\subsubsection{Normalized Tree Similarity}
We used the Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) to be able to compare the two metrics for any $n$ taxa. This index allows to scale the value of any metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are strictly identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topological recovery: a high score of this index (i.e. towards 1) means that the topology is highly conserved between the two trees; on the opposite, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as much as expected when comparing two random trees.

\subsubsection{Tree comparisons}
For the Maximum Likelihood and Bayesian consensus trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Robinson-Foulds and Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012}. For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). % TG: Maybe remove supplementary here
We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}.

@@

For each of the 125 "missing-data" tree, we obtained Robinson-Foulds and Triplets distance distributions from either 50 or 50*1000 pairwise comparisons. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. Finally we tested for significant differences among the different distributions of trees using non-parametric group and pairwise comparisons by using respectively Kruskal-Wallis and Nemenyi-Damico-Wolfe-Dunn test (CITE R PACKAGE)\citet{ruxtontime2008}.

@@

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/colour/Parameters-AllMethods-RF+Tr-small.pdf}
\caption{Comparison between the effect of missing data and the tree inference method on topological recovery. The amount of missing data for each parameter is represented on the x axis. The topological recovery is represented on the y axis, both using Robinson-Foulds distance (upper row) and Triplets distance (lower row). Points represent the modal value of each distribution ; thick solid and thin dashed lines represents respectively the 50\% and 95\% confidence intervals or the distributions. The Maximum Likelihood trees are represented in black, the Bayesian consensus trees in red, the bootstrap trees in green and the posterior tree distribution in blue.}
\label{Fig_Results_1} %Differences within a subset of parameters and between methods
\end{figure}

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/BW/Global-ML+consensus-RF+Tr-small.pdf}
\caption{Trend of the effect of missing data on topological recovery on ML and consensus trees. The amount of missing data per parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is represented along the x axis. The colour gradient from white to black represents respectively, 0\%, 10\%, 25\%, 50\% and 75\% of missing data. The topological recovery is represented on the y axis, both using Robinson-Foulds distance (upper row) and Triplets distance (lower row). Points represent the modal value of each distribution ; thick solid and thin dashed lines represents respectively the 50\% and 95\% confidence intervals or the distributions. The Maximum Likelihood trees are represented in black and the Bayesian consensus trees in grey.}
\label{Fig_Results_2} %Differences between all the parameters and between two methods (ML vs BC)
\end{figure}

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/colour/PairComp_within-Baycon-RF+Tr.pdf}
\caption{Pairwise Bhattacharyya Coefficients within the Bayesian consensus trees. The pairwise trees comparisons are represent on both axis. The colour gradient from white to black represents respectively, 0\%, 10\%, 25\%, 50\% and 75\% of missing data. The matrix represents the values of pairwise Bhattacharyya Coefficients going from green (BC=1) to red (BC=0). A. Results for the Normalised Robinson-Foulds distance. B. Results with the for Normalised Triplets distance.}
\label{Fig_Results_3}
\end{figure} %Pairwise BC for the Bayesian consensus trees for RF and TR. (parameters differences)

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/colour/PairComp_without-RF+Tr.pdf}
\caption{Distribution of the Bhattacharyya Coefficients between methods. Curves represents the kernel density estimations of the BC between the four different methods. A. Results for the Normalised Robinson-Foulds distance. B. Results with the for Normalised Triplets distance. Pikes around the two different extremes values of the BC show the two pairs of most dissimilar methods (Bayesian consensus vs. Bootstraps and Bayesian consensus vs. Bayesian tree posterior distribution) for the Normalised Robinson-Foulds distance ; and the most similar methods (Bootstraps and Bayesian tree posterior distribution) for the Normalised Triples methods.}
\label{Fig_Results_4} %Comparisons between methods, both in RF and Triplets.(method differences)
\end{figure}

\subsection{Effect of the method on topological recovery}
-With RF Bayesian significantly better than ML for clade conservation.

-With RF Bayesian consensus trees plateaus around 0.8 topological recovery at worst whatever combination of missing data.

-With Tr however, ML is more performant but not significantly. When looking at uncertainty, both results tend to just overlap.

\subsection{Effect of the missing data on topological recovery}
-No big difference between the parameters individually

\subsection{Effect of the amount of missing data}

@@

\subsection{Building the trees}
Simulating evolutionary history matrices still remains a big drawback in theoretical phylogenetics. %Hahahaha.... feck.
The size of our simulated matrices was at least two orders orders of magnitude lower than usual matrices, both for the molecular part \citep[e.g.][]{springermacroevolutionary2012} and the morphological part \citep[e.g.][]{nithe2013}. % NC: Caveats later on
This configuration probably lead to globally low phylogenetic signal as well as the intrinsic difficulties to simulated characters with phylogenetic signal.
Even though molecular characters evolution (and therefore simulation) only depends of a small number of parameters (i.e. the base frequencies and the substitution matrix), simulating molecular matrix with a strong evolutionary signal is still complicated when generating unrealistically small matrices.
For morphological characters, the underlying pattern of their evolution are often more complex and ruled by more parameters than molecular characters (i.e the number of character states, the states frequencies, the substitution matrix and the statistical model used) \citep{Pagel22011994,wagner2000,lewisa2001}.

Also morphological characters studies involve many potential statistical pitfalls (e.g. independent characters violation, rate variation - \citet{davalosintegrating2014}) and especially (i) incongruence with molecular signal and (ii) homoplasy. %REF probably April has something about it
\begin{enumerate}
\item{First, morphological data can display a different signal than molecular data, especially in small matrices. %REF
This might lead to a controversial phylogenetic signal in the overall matrix and lower down the support values.
However, regarding empirical data studies, most of the groups shows fairly congruent morphological and molecular phylogenetic signal \citep[e.g.][]{leerates2013}.}
%The main difference between molecular and morphological characters is that, because of historical and practical reasons, a morphological character matrix do no contains autapomorphies and invariant characters which are usually more common than synampomorphies in molecular character matrices (cite).
\item{Secondly, in this study, we made the assumption that theoretically, morphological characters are randomly distributed on an organism however it seems clear that empirical morphological data does not act randomly \citep{sansomfossilization2013}.
However, following our simulation assumption of random character distributions, if they accumulate through time in the same way as the majority of the molecular characters then, homoplasic characters are expected to appear randomly through time \citep{davalosintegrating2014}.
Therefore, homoplasy is expected to be more important (by chance) in bigger morphological matrices \citep{davalosintegrating2014}.
After a reaching a critical amount of morphological characters, adding new ones increases homoplasy \citep{wagner2000}.}
\end{enumerate}
Our simulation parameters both decrease the phylogenetic signal (difficulty to simulate morphological characters and size of the matrix) as well as it increases it (reduction of homoplasy in the morphological part of the matrix).
Therefore, these drawbacks seems to have only a minor impact on the main results of this study: the incapacity of recovering any topology in Bayesian inference.

\subsection{Comparing topologies} % NC: Is any of this really relevant to the central questions of the paper? I don't think so... Maybe some of this could go in the methods where you chose these metrics. But it may not even be needed there.
Comparing topologies is a crucial question in phylogenetics but has always been hard to normalise because of the vast amount of different metrics used as proxies for different aspects of tree similarity/dissimilarity \citep{agapowpower2002}.
Because our global framework is to study how to include efficiently fossils into phylogenies, we chose metrics reflecting the most interesting aspect of this global question: where do individual taxa (i.e. the fossils) branch in the tree.
The Robinson-Foulds \citep{RF1981} and the Triplets \citep{critchlowthe1996} metrics are more sensible to taxa and clade placement than other tree comparison metrics \citep[e.g.][Imbalance metric]{kirkpatricksearching1993} and where therefore favored.
Also by using the Normalized Tree Similarity index \citep{Bogdanowicz2012} we emphasize the aspect of "good" phylogenetic signal \textit{versus} random phylogenetic signal because it normalized the values of the metric by correcting for the expected value when comparing random trees (NTS = 0).

The Robinson-Foulds metric is a conservative tree topology metric, it is more sensible to single taxon displacement because it will count clades as similar only if they are composed of the same number of taxa with the same topologies \citep{RF1981}.
When getting closer to the root of the tree, displacement of single taxon makes the clades not being exactly identical any more even if the clade still contains all the other taxa in both trees.
On the other hand, the Triplets method is measuring the position of each taxon towards to other reference taxa \citep{critchlowthe1996}.
It will penalise only trees where taxa get removed furthest from their original clade.
Regarding our problematic (how does missing data influence topological recovery in trees containing both living and fossil taxa) we are more interested the placement of taxa (i.e. where does the fossil branch) than the exact conservation of clades.

%Need a transition!

Although the idea of comparing two single trees is straightforward, it becomes more complex when comparing trees distributions (e.g. Bayesian posterior distributions).
The introduction of our Random Pairwise Bayesian Tree Comparison methods allows to summarize the comparison of two trees distributions by picking up the most frequent signal in the distribution of the pairwise comparisons (i.e. the mode).
Even though this method is subject to randomness and might artificially increase or decrease the score of the studied metric, simulations showed that when a sufficient amount of random comparisons is performed, this method doesn't seem to be subject to randomness (e.g. 1000 random pairwise comparisons - see ~\ref{supplementaries}).

\subsection{Maximum Likelihood versus Bayesian}
The main results of our analysis shows that Bayesian inference fails to recover any Topology in a Total Evidence method framework (Fig. ~\ref{Fig_Results}).
This results is surprising regarding the behaviour of Maximum Likelihood inference (Fig. ~\ref{Fig_Results}) as well as the Bayesian inferences performed on empirical data \citep{ronquista2012,schragocombining2013}.
However, it is important to note that this effect was not mentioned in the aforementioned empirical tests because both studies used a Bayesian approach with fixed topology \citep{ronquista2012,schragocombining2013}.

In our case, we suspect that this inability of Bayesian methods to recover topology is due to the intrinsic structure of a Total Evidence matrix.
In fact, as previously studies have shown, missing data doesn't seem to be a major drawback as long as the missing data is randomly distributed ~\citep[e.g.][]{wiensmissing2003,rouresite-specific2011,sansomfossilization2013}.
However, in Total Evidence matrices, the majority of the missing data is not randomly distributed but concentrated in the molecular part of the matrix for the fossil taxa (i.e. the missing-data sub-matrix).
This leads to high decrease of topological recovery when using non optimal criterion approach (i.e. Bayesian inference) because of the high variance in the near-likeliest solutions sampled in the Bayesian posterior distribution.
In opposition, when applying optimal criterion approach (i.e. Maximum Likelihood), it is still possible to sample the likeliest tree. %probably need to develop that way more, especially with actually analysing the likelihood landscape.

>>>>

-Overall missing data is doesn't affect topology in Bayesian (around minimum 70\% of topological topology in the worth scenario).

@@

The effect of missing data on phylogenetic inferences has been widely studied \citep{wiensmissing2003,wiensmissing2006,wiensmissing2008,lemmonthe2009,rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,wrightbayesian2014}. Missing molecular data has been seen by some authors as an issue because it can, in some part of the tree, decrease phylogenetic signal (i.e. the evolutionary information contained within the matrix allowing to infer topology and branch length)% NC: define phylogenetic signal here. 
    , especially when using large matrices \citep{lemmonthe2009}. However, this may not be a major issue because phylogenetic signal is easily increased by:
(i) including a "modest" number of highly-covered genes (i.e. approximatively of the genes that are coded for most of the species; \citealt{rouresite-specific2011}) %NC: what do you mean by highly-covered?
(ii) adding a greater number of taxa (especially slowly-evolving taxa or taxa close to the outgroup; \citealt{rouresite-specific2011}); and (iii) choosing more appropriate models of sequence evolution \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Similarly, missing morphological data might be seen as either a major or minor issue for accurately inferring phylogenies depending on the study in question \citep{wiensmissing2003,sansomfossilization2013,pattinsonphylogeny2014}. Because soft-tissue characters are rarely preserved in the fossil record, missing data is mainly found in these characters, and is therefore not randomly distributed which can lead to biased placement of fossil taxa in phylogenies (e.g. \citeauthor{sansomfossilization2013} \citeyear{sansomfossilization2013} but see \citeauthor{pattinsonphylogeny2014} \citeyear{pattinsonphylogeny2014}). However, the phylogenetic signal is not related to the amount of missing data \textit{per se} but to the number of informative characters for each taxon, therefore missing data is less of an issue than the number of shared informative characters \citep{wiensmissing2003}.
% NC: I'm a little dubious of what this paragraph is trying to get at in places. We should discuss at some point. TG: my idea was to do the "literature-review" part of the intro: missing data is already widely studied (answer to the comment: "hun... We already knew that") but it has never been studied in TEM framework. And that's kind of the point we're getting toward with our work: we don't really care about the effect of missing data per se (i.e. what's the threshold were it starts fucking up the phylogeny) but more the fact that Bayesian fails.

@@

We inferred the morphological matrix using the R package ape v3.0-11 \citep{paradisape:2004} to generate a matrix of 100 character sites for 51 taxa. We assigned the number of character states (either two or three) for each morphological character by sampling with a probability of 0.85 for two states characters and 0.15 for three state characters. These probabilities were selected using the overall distribution of character states extracted from 100 published empirical morphological matrices (See \hyperref[SupplementaryMaterial]{Supplementary Material Section 1}). We then ran an independent discrete character simulation for each character using the "true" tree with the character's randomly selected number of states (two or three) and assuming an equal rate of change (i.e. evolutionary rate) from one character state to an other \citep{Pagel22011994}. This method allows us to have only two parameters for each character: the number of states and the evolutionary rate. For each character, the evolutionary rate was sampled from a gamma distribution with $\alpha$ = 0.5. We used low evolutionary rate parameters (i.e. $\alpha$) to avoid homoplasy in the morphological part of the matrix and create a clear phylogenetic signal \citep{wagner2000,davalosintegrating2014}.

@@

\subsubsection{Normalized Tree Similarity}
We used the Normalized Tree Similarity index, $NTS_m$ (\citealp{Bogdanowicz2012}) to be able to compare the two metrics for any $n$ taxa. This index allows to scale the value of any metric $m$ (either Robinson-Foulds or Triplets distance in our study) to the expected value of the metric $m$ when comparing two random trees (see \hyperref[SupplementaryMaterial]{Supplementary Material S2}). When $NTS_m$=1, the two trees are strictly identical, when $NTS_m$=0 the trees are no more different than expected when comparing two random trees and when $NTS_m$$<$0, the difference between the two trees is greater than when comparing two random trees. In our study we used the $NTS_m$ index as a proxy for topology: a high score of this index (i.e. towards 1) means that the topology is highly conserved between the two trees; on the opposite, a low score of this index (i.e. towards 0) means that the topological difference between the two trees is as much as expected when comparing two random trees.

\subsubsection{Tree comparisons}
\label{tree_comparisons}
For the Maximum Likelihood and Bayesian consensus trees we performed pairwise comparisons between the "best" tree and each "missing-data" tree using both the Robinson-Foulds and Triplets metrics with the TreeCmp java script \citep{Bogdanowicz2012}. For each metric, we then normalized the value using the Normalized Tree Similarity scaled by the mean value of 1000 pairwise random tree comparisons for the metric in question and n = 51 taxa (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 2}). % TG: Maybe remove supplementary here
We compared each "missing-data" tree with the "best" tree for each of our 50 simulation runs resulting in 50 comparisons for each "missing-data" tree. We calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}.

@@

For each of the 125 "missing-data" tree, we obtained Robinson-Foulds and Triplets distance distributions from either 50 or 50*1000 pairwise comparisons. We then calculated the mode and the 50\% and 95\% confidence intervals from the resulting distribution using the hdrcde R package v3.1 \citep{hdrcde}. %Finally we tested for significant differences among the different distributions of trees using non-parametric group and pairwise comparisons by using respectively Kruskal-Wallis and Nemenyi-Damico-Wolfe-Dunn test (CITE R PACKAGE)\citet{ruxtontime2008}.

In order to investigate the effect of the parameters and/or the methods used in this simulations, we measured the similarity among the different distributions using the Bhattacharyya Coefficient \citep{Bhattacharyya}. The Bhattacharyya Coefficient is the probability of overlap between two distributions, ranging from 0 to 1 \citep{Bhattacharyya}. For each method (Maximym likelihood trees, Bayesian consensus trees, Bootstraps and Bayesian posterior trees) and metric (Robinson-Foulds and Triplets distance), we used this coefficient in two ways in order to:
\begin{enumerate}
\item assess the effect of the method on all parameters:

We performed a pairwise comparison of each parameter between each pair of methods. This resulted in 125 comparisons per method and per metric (each parameter combination within one method compared to the same parameter combination in the other method). We then compared the distribution of these pairwise comparisons to see the global difference between two methods: if the method have really similar results, we would expect the distribution to be cluster around 1 (high probability of overlap between the distributions) and if the methods are really dissimilar around 0 (low probability of overlap - Fig.~\ref{Fig_Results-paircomp_without}).
\item assess the effect of the parameters within a method:

We performed a pairwise comparison per method and per metric for each combination of parameters. This resulted in 7875 pairwise comparisons per method and per metric (triangle of a 125*125 matrix). We represented these results as a triangular matrix with the values of each pairwise comparison coloured according to the value of the Batthacharyya Coefficient (coloured in green when the distributions overlap completely and in red when they don't - Fig.~\ref{Fig_Results-paircomp_within}).
\end{enumerate}

@@

\subsection{Effect of missing data on topology}

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/colour/Parameters-AllMethods-RF+Tr-small.pdf}
\caption{Comparison between the effect of missing data and the tree inference method on topology. The amount of missing data for each parameter is represented on the x axis. The topology is represented on the y axis, both using Robinson-Foulds distance (upper row) and Triplets distance (lower row). Points represent the modal value of each distribution ; thick solid and thin dashed lines represents respectively the 50\% and 95\% confidence intervals or the distributions. The Maximum Likelihood trees are represented in black, the Bayesian consensus trees in red, the bootstrap trees in green and the posterior tree distribution in blue.}
\label{Fig_Results-permeth_perparam} %Differences within a subset of parameters and between methods
\end{figure}

As it would be expected form the literature, the amount of missing data in the morphological matrix does decrease the ability to recover the right topology, regardless of the parameter, the method or the metric used \citep{rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,wrightbayesian2014}. However each variable does not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam}). Regarding the conservation of entire clades (i.e. the Ronbinson-Foulds distance), the Bayesian consensus trees outperform the other methods and the amount of missing data in the living taxa ($M_{L}$) decreases the most rapidly the clade conservation when using this method. However, when looking at the position of wild card taxa (i.e. the Triplets distance), the Maximum Likelihood outperforms the other methods but with wider distributions overlap (Fig.~\ref{Fig_Results-paircomp_without}-B).

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/BW/Global-ML+consensus-RF+Tr-small.pdf}
\caption{Trend of the effect of missing data on topology on ML and consensus trees. The amount of missing data per parameter ($M_{L}$, $M_{F}$ and $M_{C}$) is represented along the x axis. The colour gradient from white to black represents respectively, 0\%, 10\%, 25\%, 50\% and 75\% of missing data. The topology is represented on the y axis, both using Robinson-Foulds distance (upper row) and Triplets distance (lower row). Points represent the modal value of each distribution ; thick solid and thin dashed lines represents respectively the 50\% and 95\% confidence intervals or the distributions. The Maximum Likelihood trees are represented in black and the Bayesian consensus trees in grey.}
\label{Fig_Results-global_perparam} %Differences between all the parameters and between two methods (ML vs BC)
\end{figure}

When looking at the global trend of the Bayesian consensus trees for all the parameters combinations, this method outperforms Maximum Likelihood for the Ronbinson-Foulds distance and plateaus to a minimal modal tree similarity of 0.69 regardless the amount of missing data (Fig.~\ref{Fig_Results-global_perparam} - see also \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). However, regarding the Triplets distance, Bayesian consensus trees seems to perform poorly but with great confidence intervals overlaps (Fig.~\ref{Fig_Results-global_perparam} and Fig.~\ref{Fig_Results-paircomp_within}). Results for the global trend comparison between the uncertainty methods (Bootstrap and Bayesian posterior tree distribution) are available in the supplementary materials (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

\subsection{Effect of the method on topology}

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/colour/PairComp_without-RF+Tr.pdf}
\caption{Distribution of the Bhattacharyya Coefficients between methods. Curves represents the kernel density estimations of the BC between the four different methods. A. Results for the Normalised Robinson-Foulds distance. B. Results with the for Normalised Triplets distance. Pikes around the two different extremes values of the BC show the two pairs of most dissimilar methods (Bayesian consensus vs. Bootstraps and Bayesian consensus vs. Bayesian posterior trees) for the Normalised Robinson-Foulds distance ; and the most similar methods (Bootstraps and Bayesian posterior trees) for the Normalised Triples methods.}
\label{Fig_Results-paircomp_without} %Comparisons between methods, both in RF and Triplets.(method differences)
\end{figure}

When looking at the comparisons between the methods using the Robinson-Foulds distance, the Bayesian consensus trees have the least overlap with respectively the Bayesian posterior trees and the Bootstrap (Fig.~\ref{Fig_Results-paircomp_without}-A). When regarding at the Triplets distance, the two method that have the most distribution are the Bootstraps and the Bayesian posterior trees (Fig.~\ref{Fig_Results-paircomp_without}-B).

\subsection{Effect of the missing data parameters on topology}

\begin{figure} 
\centering
    \includegraphics[width=1\textwidth]{Figures/colour/PairComp_within-Baycon-RF+Tr.pdf}
\caption{Pairwise Bhattacharyya Coefficients within the Bayesian consensus trees. The pairwise trees comparisons are represent on both axis. The colour gradient from white to black represents respectively, 0\%, 10\%, 25\%, 50\% and 75\% of missing data. The matrix represents the values of pairwise Bhattacharyya Coefficients going from green (BC=1) to red (BC=0). A. Results for the Normalised Robinson-Foulds distance. B. Results with the for Normalised Triplets distance.}
\label{Fig_Results-paircomp_within}
\end{figure} %Pairwise BC for the Bayesian consensus trees for RF and TR. (parameters differences)

When looking at the pairwise Bhattacharyya Coefficients between all the parameters combinations for Bayesian consensus trees, the number of missing living taxa ($M_{L}$) results the lowest probabilities of overlap (right lower corner in Fig.~\ref{Fig_Results-paircomp_within}-A). However, when looking at the Triplets distance, because of the high overlap in confidence intervals, there is no major region with low probability of overlap. One should note however that the bottom line of the matrix where the Bhattacharyya Coefficients are very low is due to comparing the distribution of the pairwise comparison of the best tree vs. the best tree, leading to a Normalised triplet score of 1 every time (with no variance, see Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). All the other pairwise comparisons are available in the supplementary materials (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

%§1
%Results: global
%   -Missing data increase decreases topological recovery
%The results of our simulations are in adequacy with recent analysis on the effect of missing data in morphological matrices \citep{rouresite-specific2011,sansomfossilization2013,pattinsonphylogeny2014,wrightbayesian2014}. Our results are showing that ability to recover the correct topology decreases with the amount of missing data, regardless the metric, the method or the parameters considered. In more details, our results shows however a clear superiority of the Bayesian consensus tree over other methods, at least when considering the Robinson-Foulds distance (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). Also, our results suggest that the missing data for missing taxa parameter ($M_{L}$) shows a slightly higher decrease in the ability to recover the correct topology (Fig.~\ref{Fig_Results-paircomp_within}).
%Results 1
%   -However, methods and metrics are unequal
%Results 2
%   -Additionally, parameters have different effect
%Wrap up results
%   -Thus, Baycon is better with loads of living

%§2
%Results from litterature
%   -Missing data per se is not a problem, problem is in the amount of overlap
%Proposed hypothesis
%   -One can reduce overlap by adding living taxa

%§3
%Discussing the metrics in detail
%   -RF is different than triplets and better.
%However, our results can be contrasted whether one looks at the Robinson-Foulds or the Triplets distance. When using the second one, the confidence interval of our different parameters combinations per methods seems higher (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}) and the probability of overlap between the distribution of each parameter is high (Bhattacharyya Coefficient close to 1 - Fig.~\ref{Fig_Results-paircomp_within}-B). The contrast between the results of the two metrics can be explained by the actual nature of the metrics:
%\begin{enumerate}
%\item{The Robinson-Foulds metric is a conservative tree topology metric, it is more sensible to single taxon displacement because it will count clades as similar only if they are composed of the same number of taxa with the same topologies \citep{RF1981}. When getting closer to the root of the tree, displacement of single taxon makes the clades not being exactly identical any more even if the clade still contains all the other taxa in both trees. This metric illustrates therefore the conservation of clades among the compared trees (i.e. when the normalised Robinson-Foulds distance is close to 1, most of the clades in the two trees are identical).}
%\item{On the other hand, the Triplets method is measuring the position of each taxon towards two other reference taxa \citep{critchlowthe1996}. It will penalise only trees where taxa get removed furthest from their original clade. This metric illustrates therefore the amount of wildcard taxa \citep{kearneyfragmentary2002} between the to compared trees (i.e. when the score is closed to 1, view wildcard taxa are present in the tree).}
%\end{enumerate}
%Therefore, in our study, both metrics are showing two different aspects of tree topology. Also, disregarding their fundamental differences in accounting for topological distance, \citet{kuhnerpractical2014} have recently proposed that the Ronbinson-Foulds distance greatly outperforms the Triples distance in signalling differences in topology.

%§4
%Discussing the results 1 in detail
%   -Why is Baycon better than the others?
%Following \citet{kuhnerpractical2014}'s conclusions on the superiority of the Robinson-Foulds distance, the best method to use for recovering the correct topology in a Total Evidence framework is using the Bayesian consensus tree (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). In fact, regardless of the amount of missing data, this method always displays the highest normalised Robinson-Foulds score (see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). When comparing this the probability of overlap in this method between the other onces, it is counter intuitive to note that the Bayesian consensus tree shows the highest dissimilarity with the Bayesian posterior trees from which the consensus tree is build (Fig.~\ref{Fig_Results-paircomp_within}-A). Note that it is also highly dissimilar to the Bootstraps. However that can be linked to the fact that both Bootstraps and Bayesian posterior trees are measured as 125*1000 random pairwise comparisons instead of just 125 tree comparisons for the Bayesian consensus trees. The fact that the 1000 trees to be compared for each of the 125 pairwise comparisons can add noise to the results for these two methods due to the random factor in the pairwise comparisons (see \hyperref[tree_comparisons]{Methods - Tree comparisons}).

%§5
%Discussing the hypothesis 2 in details
%   -Which parameters does what?
%Which parameter does what: for $M_L$, one should put more effort in using natural museums history collections for coding the missing morphological data for living taxa if possible %go to the museum!; for the $M_F$ parameter, the amount of missing data unfortunately depends on the quality of the fossil record and can not be actively improved and depends on exceptional discoveries \citep[e.g.][]{nithe2013}; finally for the $M_C$ parameter, improvement can be done by vast collaborative projects in order to gather as much characters as possible \citep[e.g.][]{O'Leary08022013}. Therefore, we advocate the importance of coding morphological characters for the most living taxa possible and with the most characters possible, potentially by using collaborative projects portals such as morphobank \citep{morphobank}.  
%   -Why is ML more important than others?
%Our results are consistent with previous works and are showing that the amount of missing data \textit{per se} does not reduce the ability to recover the correct topology \citep[e.g.]{kearneyfragmentary2002,wiensmissing2003,pattinsonphylogeny2014}. In fact, as shown previously, the amount of overlap in the matrix is more crucial \citep[e.g.]{rouresite-specific2011}. In our case, the overlap in data is greatly reduced when different parameters are combined with a high amount of missing data (e.g. $M_{L}$=75\%, $M_{F}$=75\% and $M_{C}$=75\% - Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). In his scenario the overall number of missing characters $M_{C}$ affects the proportion of overlapping data by reducing the overall size of the matrix (Fig.~\ref{Fig_Results-global_perparam}). However it is important to note that in the pairwise Bhattacharyya Coefficient analysis, the area with the lowest probability of distribution overlap is led by comparing the trees with no missing data for the living taxa to the trees with 75\% missing data for the living data (Right lower corner in Fig.~\ref{Fig_Results-paircomp_within}).
